# R Refresher Part 1: First Movements

## Created by Mauricio Cantor, with modifications by Laura J. Feyrer, Ana Eguiguren, and Reid Steele

This section covers basic R workspace management.

*NOTE:* On refresher pages, some code lines will be commented out to avoid file structure issues surrounding saving files, downloading packages, and changing working directories.

## Working Directories:

Your working directory is the default file path your R session is working under. You can find your working directory using the `getwd()` command. If you're working in an R project, it will be set to the path of the project by default.

```{r}
# First of all, let's check our working directory 
# (where we get files from and save our work in)
getwd()
```

You can change your working directory if needed using `setwd()`.

```{r}
# you can change it using the function 'setwd()' and 
# pasting the directory path in quotes inside the parentheses
# note that in R we always use "/" or "\\" instead of "\" on mswindows systems

# For example:
# setwd("C:/Users/R3686/OneDrive - Dalhousie University/Documents/Teaching/Anaylsis of Biological Data/Refresher/")

# You can also set up your working directory in Rstudio 
#by clicking in Session>Set Working Directory>Choose Directory; 
#or in R by clicking in File>Change Directory

```

You can see what is inside your working directory using `dir()`, and pull out specific files of interest using a pattern with `list.files()`

```{r}
# Let's take a look in what we have in our working directory
dir()

# Pick out all the .csv files
list.files(pattern = '.csv')

# If you are already sick of R, you can quit by typing:
#q()

```

## Workspace Management

You can save your R workspace into your current directory using `save.image()`, or save specific objects using `save()`. This is useful for saving results that might change due to randomization, or that may take a long time to generate, such as complex models. You can then load these objects back in later using `load()`.

```{r}

# You can save your work before you go (useful sometimes)
# save.image('Workspace.RData')#but remember to give it a useful name
# save(x, 'Workspace.RData')
# 
# load('Workspace.RData')

```

Your R workspace is the environment which contains all of the R objects you've made. By default, everything in your R workspace is visible in the top-right panel in RStudio. You can list everything in your workspace using `ls()`, and delete individual objects using `rm()`. These commands can be combined using `rm(list=ls())` to completely clear your workspace.

You can make objects in your workspace using `<-` or `=`. Object names can be anything - try to make them informative so that you'll know what they are. Remember, everything is case-sensitive in R. X is a different object from x.

```{r, error = T}

# Let's take a look in our workspace
ls()

# It's empty: no objects.Sure? what about x?
x
# See, there's no such object. Let's create one called x
x <- 2

x = 2

# '<-' assigns values to objects, works the same as '='
# here is our object x

x


# Let's take another look in our workspace. See all you objects there?
ls()

# RS Note: We can use ls to clear the entire workspace
rm(list = ls())
```
